\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-24 paper/chance_memo.tex
misc edits, almost ready for resubmission

  The $t$-value for the quadratic term is -1.88 and the corresponding $p$-value is .062, which is not
  particularly compelling.  We have accordingly dropped the quadratic
  terms in the scatter plots on p.\ 4 and, implicitly, from the
  Studentized residual plots on p.\ 6.
  %% Note: xr labels are
  %% breaking; looks like some sort of
  %% label clash with another package.
  %% Hardcoding here is poor style.

@@

  We have added age to our simulation model, and the results are
  qualitatively equivalent.  In addition, if one adds age to the
  linear specification depicted in our basic scatter plot (now without
  a quadratic term, as noted above), age is a significant predictor of
  marathon finishing time, even controlling for personal best
  ($t \approx 2.426$, $p \approx 0.017$).  Older runners are slower, \emph{ceteris
    paribus}.

@@

@@

  We have highlighted the left side bins in red and annotated the plot 
  with a description of the size of the those bins. This emphasizes the finishes
  that are the main interest of the result.

>>>>

  The $t$-value for the quadratic term is -1.88 and the corresponding
  $p$-value is 0.062, which is not particularly compelling.  We have
  accordingly dropped the quadratic terms in the scatter plots on p.\
  4 and, implicitly, from the Studentized residual plots on p.\ 6.  In
  terms of the latter suggestion about, Figure 1b contains a plot of
  half marathon split times versus actual finishing times.
  %% Note: xr labels are
  %% breaking; looks like some sort of
  %% label clash with another package.
  %% Hardcoding here is poor style.

@@

  We have added age to our simulation model, and the results are
  qualitatively equivalent.  In addition, if one adds age to the
  linear specification depicted in our basic scatter plot (now without
  a quadratic term, as noted above), age is a significant predictor of
  marathon finishing time, even controlling for personal best
  ($t \approx 2.426$, $p \approx 0.017$).  Older runners are slower,
  all things equal.

@@

  We agree with the thrust of this comment, and our manuscript now
  emphasizes the lack of independence.  Insofar as dependence in
  sampling typically diminishes variance, our lack of independence in
  the pair plots should make our results conservative. This means that
  that the observed Hahner and Kim compression in final results,
  conditional on prior best marathon times, is even more unusual.

@@

  We agree with this suggestion.  Accordingly, in the histograms that
  describe simulation results, we have highlighted left side bins in
  red and annotated the plots with descriptions of the sizes of key
  areas.  This emphasizes the finishes that are most important for our
  results.

### 2017-02-24 paper/paper.tex
misc edits, almost ready for resubmission

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; split times from the Rio marathon course at 5 kilometers, 10
kilometers, and so forth; and, finishing time. We cannot directly
observe the effort that an individual put into the race, and we do not
know why some runners have DNF results.  Some runners may have injured
themselves on the course and accordingly dropped out, and others may
have dropped out, uninjured, in anticipation of an unsatisfactory
result. Of the 156 marathon starters, 133 completed the race and 23
DNFed at various locations throughout the course. The overall DNF rate
was thus $\frac{23}{156} \approx 0.15$, and the relatively small
sample size at our disposal means that a 95\% confidence for this rate
is fairly wide, namely, $\left(0.098, 0.22\right)$.

Kurschilgen's accusation against the Hahner twins has two components,
that these two women ran slowly \emph{and} that they finished
simultaneously.  We suspect that Kurschilgen would not have expressed
ire at the Hahners had they finished in 1st and 2nd place in Rio,
hand-in-hand with wide grins.  Thus, our investigation of the charges
that Kurschilgen offered will distinguish between a slow finish versus
a simultaneous finish.

@@

One way that we might assess whether an athlete's Rio finish was
unusual---or, say, whether two athletes' finishes were jointly
unusual---is by comparing a runner's observed Olympic women's marathon
finishing time on August 14, 2016, with a measure of her underlying
marathon talent.  For the latter we use prior personal best marathon
times.  This is a natural measure of marathon talent, we believe, but
it is not entirely free of complications.  For example, personal best
times are potentially confounded by the marathon courses on which they
were set; some courses, like the Berlin marathon, are known for
relatively fast times.  In addition, personal best times may be
confounded by conditions, like weather, that varied across races.
Finally, personal best times may not capture Olympic raceday
idiosyncrasies that could affect individual runners.  For example, a
runner could have woken up with a minor cold on the morning of August
14, 2016.  With these concerns in mind, we use a runner's half
marathon split time from the Olympic marathon as a secondary measure
of athletic skill in marathoning.  Our two measures of marathon talent
are featured in the visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain %second-order polynomial regression lines. 
a dashed line representing the line of best fit.
The ticks along the
horizontal axes in both plots indicate times, either personal bests or
half marathon splits, of runners who earned DNF
results.\label{scatterplots}

@@

Another perspective on the extent to which the Hahners' overall
marathon finishing times were not unusual can be gleaned from Figure
\ref{fig:residuals}.  This figure presents cumulative distributions of
the polynomial regression residuals from the regression models
displayed in the two panels of Figure \ref{fig:scatter}.  In
particular, we calculated the residuals from Figures
\ref{fig:45degreeplot} and \ref{fig:45degreeplot_half}, Studentized
them, and then arranged resulting Studentized residuals from least to
most along the horizontal axes in Figures
\ref{fig:studentizedresiduals} and
\ref{fig:studentizedresiduals_half}.\label{residualplots}

@@

The cumulative residuals depicted in the two panels of Figure
\ref{fig:residuals} are consistent with our interpretation of the
Hahner twin's Olympic finishing times as not particularly
remarkable. With personal best time as a measure of marathon talent as
in Figure \ref{fig:studentizedresiduals}, the two Hahner residuals are
in the right tail of the residual distribution; however, their
locations are not extreme.  One residual is located at the 88th
percentile and the other at the 93rd.  While these two residuals are
in the tail end of the residual distribution, they are not the major
outliers that would lead us to think that the Hahner twin's marathon
finishing times were extremely slow.  Moreover, the residuals for the
Kim twins are similarly not remarkable, although they are in the left
tail of the residual distribution, indicating that the Kims ran faster
than one would have expected.  If one uses half marathon split times
as a measure of marathon talent, as in Figure
\ref{fig:studentizedresiduals_half}, a similar conclusion is reached.
Neither Hahner twin had a finishing time that was particularly unusual
given her half marathon split, and this applies to the Kim twins as
well.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's black dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

@@

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to subsequent DNFs.  The usual color scheme applies here,
and the Estonian triplet DNF prior to the half marathon split is
evident. As noted in the figure, Lisa Hahner was ahead of her sister
through 15 kilometers at which point Anna took the lead.  Figure
\ref{fig:secondsbehind} contains two red dots representing the North
Korean Kim twins, but this is not visually apparent because the Kim
twins had identical split times during the entire marathon.

@@

@@

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a linear function of the runner's marathon
skill $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal
distribution. Taking each runner's personal best marathon time as a
proxy for her underlying marathon skill, we estimate $\beta_1$,
$\beta_2$, $\beta_3$ and $\sigma$ using ordinary least squares on all
finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
triplets from the sample so that that our estimates are not affected
by the twin/triplet finishes which, theoretically, could reflect the
results of coordination.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, respectively, from a multivariate normal
  distribution with mean equal to the previously estimated coefficient
  vector and covariance equal to the previously estimated covariance
  matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
\item Predict each runner's final result by combining the randomly
  generated beta coefficients and individual error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +  \tilde{\beta}_2\,Z_i + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

As these steps illustrate, our simulation repeatedly draws random
coefficient vectors, and this captures variability in what we know
about the relationship between runner talent and age and runner finish
times.  In addition, the simulation draws random disturbances for each
runner, conditional on our original estimate of regression variance;
these disturbances capture variability in runner finishing times,
notwithstanding underlying runner talent.  Importantly, the
disturbances that we draw are independent across runners.  This means
that the finishing order among runners for each simulated race will
vary.

From our simulations we can generate intervals which describe the
extent of the variability in marathon finishing times. For example, in
95\% of the simulations in which Anna Hahner completed the marathon,
her finishing time was between 8,369 seconds and 9,999 seconds. This
is consistent with Anna's observed finishing time in Rio, which was
9,932 seconds. In fact, our simulations estimate that Anna should have
finished the Rio marathon in 9,181 seconds, which is slightly more
than 12 minutes faster than her actual time. Lisa
Hahner's corresponding 95\% interval is 8,507 to 10,143 with an
observed finishing time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times. And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets). According
to our simulation, then, both Hahner twins ran appreciably slowly,
conditional on personal best times prior to the Rio Olympics and age.

@@

With an eye on the matter of simultaneous finishing, Figure
\ref{fig:simdiff} contains two histograms based on our simulated race
results. Figure \ref{fig:simulatedfinishtimes} is a histogram which
shows the distribution of absolute differences in Hahner twin
finishing time where differences are grouped in 30 second bins; counts
for the various bins are denoted by the vertical lengths of the bars.
Figure \ref{fig:simulatedranks} is similar but depicts the
distribution of the absolute differences in Hahner twin rankings.
Differences are grouped as single units ranging from no runner between
the twins to nearly 120 runners between them.

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on personal best times}
  \label{fig:simdiff}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_with_age.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_with_age.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

The histograms in Figure \ref{fig:simdiff} raise questions about the
credibility of Anna and Lisa Hahner's story and in particular suggest
that a simultaneous finish in the Rio marathon would be very rare if
Anna and Lisa had run independently of each other. For example, in
fewer than 300 of the 10,000 simulated races did Anna and Lisa Hahner
finish within 30 seconds of one another. Moreover, they finished in
consecutive rank in fewer than 200 of the 10,000 simulated races.  The
close finish that was observed, where Anna and Lisa crossed the finish
line one after the other, would have been highly unlikely if the two
German twins had raced independently of each other.

Parallel with our prior analyses, we repeated our simulations but this
time used half marathon splits as the predicting variable in our
simulations. The results of this are in Figure \ref{fig:simdiffhalf}.
While this use of marathon splits reduces the variation in the
predicted outcome of each runner and therefore reduces the expected
distance between Anna and Lisa Hahner, even with half marathon split
as a measure of ability it is still quite rare for the twins to finish
simultaneously or consecutively.

\begin{figure}[!ht]
  \centering
  \caption{Distribution of simulated results for the Hahner twins
    based on half marathon splits}
  \label{fig:simdiffhalf} 
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_half_with_age.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes_half}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half_with_age.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks_half}
  \end{subfigure}
\end{figure}

An important point regarding our simulations is the way that we
handled DNFs. As our description above indicates, we assumed that DNF
probabilities are the same for all runners and that the likelihood of
a DNF is a not function of a runner's anticipated marathon finishing
time. The rug marks in Figure \ref{fig:45degreeplot} suggest that
runners with better personal best times may be more likely to DNF than
other runners, all things equal. We suspect that this occurs because
some better runners may expend excessive energy trying to achieve a
good result in the marathon and in so doing injure or exhaust
themselves; lesser runners, in contrast, may be content to finish
respectably. Regardless of the validity of this conjecture, Figure
\ref{fig:45degreeplot} shows that the Hahner twins are representative
of the sort of runners who DNFed in Rio. Since our simulated Hahner
statistics are conditioned on both Hahner twins finishing, it follows
that they are conservative. The fact that both women finished the
marathon was notable in and of itself, and by discounting the
possibility of a Hahner DNF we are giving the benefit of the doubt to
the Hahner twins.

>>>>

For each participant who started the women's Olympic marathon, we know
several things: personal best marathon time prior to the 2016 Olympic
games; age; split times from the Rio marathon course at 5 kilometers,
10 kilometers, and so forth; and, overall finishing time. We cannot
directly observe the effort that an individual put into the race, and
we do not know why some runners have DNF results.  Some runners may
have injured themselves on the course and accordingly dropped out, and
others may have dropped out, uninjured, in anticipation of an
unsatisfactory result. Of the 156 marathon starters, 133 completed the
race and 23 DNFed at various locations throughout the course. The
overall DNF rate was thus $\frac{23}{156} \approx 0.15$, and the
relatively small sample size at our disposal yields a relatively wide
95\% confidence for this rate, namely, $\left(0.098, 0.22\right)$.

Kurschilgen's accusation against the Hahner twins has two components,
that these two women ran slowly \emph{and} that they finished
simultaneously.  We suspect that Kurschilgen would not have expressed
ire at the Hahners had they finished in 1st and 2nd place in Rio,
hand-in-hand with wide grins.  Thus, our investigation of the charges
that Kurschilgen offered distinguishes between a slow finish and a
simultaneous finish.

@@

One way that we might assess whether a women marathoner's Rio finish
was unusual---or, say, whether two finishes were jointly unusual---is
by comparing a runner's observed Olympic finishing time on August 14,
2016, with a measure of her underlying marathon talent.  For the
latter we use prior personal best marathon times.  This is a natural
measure of marathon talent, we believe, but it is not entirely free of
complications.  For example, personal best times are potentially
confounded by the marathon courses on which they were set; some
courses, like the Berlin marathon, are known for relatively fast
times.  In addition, personal best times may be confounded by
conditions, like weather, that varied across races.  Finally, personal
best times may not capture Olympic raceday idiosyncrasies that could
affect individual runners.  For example, a runner could have woken up
with a minor cold on the morning of August 14, 2016.  With these
concerns in mind, we use a runner's half marathon split time from the
Olympic marathon as a secondary measure of athletic skill in
marathoning.  Our two measures of marathon talent are featured in the
visualizations below.

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain %second-order polynomial regression lines.
dashed lines representing least squares regression fits.  The ticks
along the horizontal axes in both plots indicate times, either
personal bests or half marathon splits, of runners who earned DNF
results.\label{scatterplots}

@@

Another perspective on the extent to which the Hahners' overall
marathon finishing times were not unusual can be gleaned from Figure
\ref{fig:residuals}.  This figure presents cumulative distributions of
the residuals from the regression models displayed in the two panels
of Figure \ref{fig:scatter}.  In particular, we calculated the
residuals from Figures \ref{fig:45degreeplot} and
\ref{fig:45degreeplot_half}, Studentized them, and then arranged
resulting Studentized residuals from least to most along the
horizontal axes in Figures \ref{fig:studentizedresiduals} and
\ref{fig:studentizedresiduals_half}.\label{residualplots}

@@

The cumulative residuals depicted in the two panels of Figure
\ref{fig:residuals} are consistent with our interpretation of the
Hahner twin's Olympic finishing times as not particularly
remarkable. With personal best time as a measure of marathon talent as
in Figure \ref{fig:studentizedresiduals}, the two Hahner residuals are
in the right tail of the residual distribution; however, their
locations are not extreme.  One residual is located at the 87th
percentile and the other at the 92rd.  While these two residuals are
in the tail end of the residual distribution, they are not major
outliers that would lead us to think that the Hahner twin's marathon
finishing times were extremely slow.  Moreover, the residuals for the
Kim twins are similarly not remarkable; these two residuals are in the
left tail of the residual distribution, indicating that the Kims ran
faster than one would have expected.  Finally, if one relies on half
marathon split times as measures of marathon talent, as in Figure
\ref{fig:studentizedresiduals_half}, similar conclusions follows.
Neither Hahner twin had a finishing time that was particularly unusual
given her half marathon split, and this applies to the Kim twins as
well.

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's black dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  The points in Figure
\ref{fig:diffdiffscatter} are not independent, but they provide a
rough sense of the dispersion in finishing time differences between
runners that one might expect conditional on differences in personal
best times.

@@

Figure \ref{fig:secondsbehind} describes each Olympic runner's status
at various split times on the marathon course.  Each dot in the
figure---color as before---depicts a recorded split and the number of
seconds each runner was behind the race leader at the time.  There are
more dots at earlier splits due to subsequent DNFs.

@@

The Estonian triplet DNF prior to the half marathon split is evident
in Figure \ref{fig:secondsbehind} , which also shows that Lisa Hahner
was ahead of her sister through 15 kilometers.  The figure contains
two red dots representing the North Korean Kim twins, but this is not
visually apparent because the Kim twins had identical split times
during the entire marathon.

@@

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a linear function of the runner's marathon
skill $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i}  + \beta_2\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal
distribution. Taking each runner's personal best marathon time as a
proxy for her underlying marathon skill, we estimate $\beta_0$,
$\beta_1$, $\beta_2$, and $\sigma$ using ordinary least squares on
finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
triplets from the sample so that that our estimates are not affected
by the twin/triplet finishes which, theoretically, could reflect
runner coordination.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$ and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, and
  $\tilde{\beta}_2$, respectively, from a multivariate normal
  distribution with mean equal to the previously estimated coefficient
  vector and covariance equal to the previously estimated covariance
  matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
\item Predict each runner's final result by combining the randomly
  generated beta coefficients and individual error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +  \tilde{\beta}_2\,Z_i + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

As these steps illustrate, our simulation repeatedly draws random
coefficient vectors, and this captures variability in what we know
about the relationship between runner talent and age and runner finish
times.  In addition, the simulation draws random disturbances for each
runner, conditional on our original estimate of regression variance;
these disturbances capture variability in runner finishing times,
notwithstanding underlying marathon talent as proxied for by prior
best finishing time.  Importantly, the disturbances that we draw are
independent across runners.  Consequently, for each simulated race the
finishing order among runners will vary.

From our simulations we generate intervals which describe the extent
of the variability in marathon finishing times. For example, in 95\%
of the simulations in which Anna Hahner completed the marathon, her
finishing time was between 8,369 seconds and 9,999 seconds. This is
consistent with Anna's observed finishing time in Rio, which was 9,932
seconds. In fact, our simulations estimate that Anna should have
finished the Rio marathon in 9,181 seconds, which is slightly more
than 12 minutes faster than her actual time. Lisa Hahner's
corresponding 95\% interval is 8,507 to 10,143 with an observed
finishing time of 9,933 seconds.

The bottom line here is that our simulated marathon finishes are
consistent with the Hahner twins' finishing times. And note that the
regression model underlying the simulations was estimated without the
Hahner twins (and same for the Kim twins and Luik triplets). According
to our simulation, then, both Hahner twins did not run appreciably
slowly, conditional on personal best times prior to the Rio Olympics
and age.

@@

With an eye on the matter of simultaneous finishing, Figure
\ref{fig:simdiff} contains two histograms based on simulated race
results. Figure \ref{fig:simulatedfinishtimes} is a histogram which
shows the distribution of absolute differences in Hahner twin
finishing time where differences are grouped in 30 second bins; counts
for the various bins are denoted by the vertical lengths of the bars.
Figure \ref{fig:simulatedranks} is similar but depicts the
distribution of the absolute differences in Hahner twin rankings.
Differences are grouped as single units ranging from no runner between
the twins to nearly 120 runners between them.

\begin{figure}[!ht]
  \centering
  \caption{Distribution of Hahner twin results in 10,000 simulated
    marathons, based on personal best times}
  \label{fig:simdiff}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_with_age.pdf}
    \caption{Differences in Hahner finishing times}
    \label{fig:simulatedfinishtimes}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_with_age.pdf}
    \caption{Differences in Hahner twin places}
    \label{fig:simulatedranks}
  \end{subfigure}
\end{figure}

The histograms in Figure \ref{fig:simdiff} raise questions about the
credibility of Anna and Lisa Hahner's story and in particular suggest
that a simultaneous finish in the Rio marathon would be very rare if
Anna and Lisa had run independently. For example, in fewer than 300 of
10,000 simulated races did Anna and Lisa Hahner finish within 30
seconds of one another, and in fewer than 600 did the Hahner twins
finish within a minute of each other.  The histogram area associated
with this result is depicted in red in Figure
\ref{fig:simulatedfinishtimes}.  Moreover, the Hahner twins finished
in consecutive rank in fewer than 200 of 10,000 simulated races; the
red zone in Figure \ref{fig:simulatedranks} presents this visually.
The close finish that was observed in Rio, where Anna and Lisa crossed
the finish line one after the other, would have been highly unlikely
if the two German twins had raced independently of each other.

Parallel with our prior analyses, we repeated our simulations using
half marathon splits as the predicting variable in our simulations;
results are in Figure \ref{fig:simdiffhalf}.  While this use of
marathon splits reduces the variation in the predicted outcome of each
runner and therefore reduces the expected distance between Anna and
Lisa Hahner, even with half marathon split as a measure of ability it
is still quite rare for the twins to finish simultaneously or
consecutively.

\begin{figure}[!ht]
  \centering
  \caption{Distribution of Hahner twin results in 10,000 simulated
    marathons, based on half marathon splits}
  \label{fig:simdiffhalf} 
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth,
    keepaspectratio]{simulated_time_half_with_age.pdf}
    \caption{Simulated differences in finishing times}
    \label{fig:simulatedfinishtimes_half}
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \includegraphics[width=\textwidth, keepaspectratio]{simulated_rank_half_with_age.pdf}
    \caption{Simulated differences in places}
    \label{fig:simulatedranks_half}
  \end{subfigure}
\end{figure}

The way that we handled DNFs in our simulations is notable. As our
description above indicates, we assumed that DNF probabilities are the
same for all runners and that the likelihood of a DNF is not a
function of a runner's anticipated marathon finishing time. The rug
marks in Figure \ref{fig:45degreeplot} suggest that runners with
better personal best times may be more likely to DNF than other
runners, all things equal. We suspect that this occurs because some
better runners may expend excessive energy trying to achieve a good
result in the marathon and in so doing injure or exhaust themselves;
lesser runners, in contrast, may be content to finish
respectably. Regardless of the validity of this conjecture, Figure
\ref{fig:45degreeplot} shows that the Hahner twins are representative
of the sort of runners who DNFed in Rio. Since our simulated Hahner
statistics are conditioned on both Hahner twins finishing, it follows
that they are conservative. The fact that both women finished the
marathon was notable in and of itself, and by discounting the
possibility of a Hahner DNF we are giving the benefit of the doubt to
the Hahner twins.
